Constructing Sailing Match Race Schedules: Round Robin Pairing Lists

### 2015-03-25 paper.tex
Model all on one page

The first thing we model, in Figure~\ref{model:skipper}, is a skipper. Each
skipper $\sigma$ has a temporal view of his schedule in the array of
constrained integer variables $\timeSlot[0]$ to $\timeSlot[{t-1}]$, and a state
view in the array of constraint integer variables $\state[0]$ to
$\state[{f-1}]$. The state of this skipper in flight $i$ is found in
$\state[i]$, and this corresponds to timeslots $\timeSlot[{i \cdot m}]$ to
$\timeSlot[{m(i+1)-1}]$.

\begin{figure}[p]
\setlength{\mathindent}{0pt}\setlength{\abovedisplayskip}{0pt}\setlength{\belowdisplayskip}{0pt}
\begin{framed}
\begin{align}
    &\forall \tau \in \{0 \nldots t-1\}: \timeSlot[{\tau}] \in \{-2 \nldots t-1\} \setminus \{\sigma\} \label{var:timeSlot} \tag{V1} \\
    &\forall i \in \{0 \nldots f-1\}: \state[{i}] \in \{\FIRST,\MID,\LAST,\BYE,\END\} \label{var:state} \tag{V2} \\
    \noalign{\smallskip \nonumber}
    &\forall i \in \{0 \nldots f\}: \label{C1} \tag{C1} \\
    &\begin{alignedat}{3}
    &\hspace*{1em} \state[{i}] = \FIRST &\ &\Leftrightarrow&&\ \timeSlot[{i \cdot m}] \geq 0 \\
    &\hspace*{1em} \state[{i}] = \MID   &\ &\Leftrightarrow&&\ \exists j \in \{i\cdot m+1 \nldots i \cdot m+m-2\} : \timeSlot[{j}] \geq 0 \\
    &\hspace*{1em} \state[{i}] = \LAST  &\ &\Leftrightarrow&&\ \timeSlot[{(i+1)m - 1}] \geq 0 \\
    &\hspace*{1em} \state[{i}] = \BYE   &\ &\Leftrightarrow&&\ \forall j \in \{i \cdot m \nldots m \cdot (i+1)-1\} : \timeSlot[j] = -1 \\
    &\hspace*{1em} \state[{i}] = \END   &\ &\Leftrightarrow&&\ \forall j \in \{i \cdot m \nldots m \cdot (i+1)-1\} : \timeSlot[j] = -2
    \end{alignedat} \nonumber \\
    &\eachOccursExactlyOnce(\timeSlot, \{0, \nldots, n-1\} \setminus \{\sigma\}) \label{C2} \tag{C2} \\
    &\validStateSequence(\state) \label{C3} \tag{C3} \\
    \noalign{\bigskip \nonumber}
    & \forall i \in \{0 \nldots f-2\}: \change[{i}] \in \{0,1\} \label{V3} \tag{V3} \\
    & \totalChanges \in \mathbb{N} \label{V4} \tag{V4} \\
    \noalign{\smallskip \nonumber}
    & \forall i \in \{0 \nldots f-2\}: \change[{i}] = 1 \Leftrightarrow \state[{i}] = \BYE \wedge \state[{i+1}] \neq \BYE \label {C4} \tag{C4} \\
    & \totalChanges = \textstyle\sum_{i=0}^{f-2} \change[{i}] \label{C5} \tag{C5}
\end{align}
\end{framed}
\caption{A skipper-oriented model. A copy of these variables and constraints is
created for each skipper $\sigma$. The number of skippers is $n$, and $m$ is
the number of matches in a flight. The number of flights is $f = \lceil
n(n-1)/m  \rceil$, and there are $t = n(n-1)/2$ matches in total.}\label{model:skipper}
\end{figure}

@@

We are now in a position to model criteria 1 (each skipper sails against every
other skipper) and optimisation criteria 10 (minimise boat changes).  In
\ref{eq13}, \ref{eq14} and \ref{eq15} we present a match perspective of the
schedule, using a two dimensional array of variables $\match$, where
$\match[{i,j}]$ is the time slot that skippers $s[{i}]$ and $s[{j}]$ meet in a
match. Only the half above the diagonal is represented, and the lower half is
made up of exactly the same variables, i.e.\ $\match[{i,j}]$ is \emph{exactly}
the same constrained variable as $\match[{j,i}]$. Constraint \ref{eq15} states
that  a match between skippers $s[{i}]$ and $s[{j}]$ takes place at time $k$
(i.e.\ $\match[{i,j}] = k$) if and only if  skipper $s[{i}]$'s $k^{th}$ time
slot is skipper $s[{j}]$ and conversely that skipper $s[{j}]$'s $k^{th}$ time
slot is skipper $s[{i}]$).  Constraints \ref{eq16}, \ref{eq17} and \ref{eq18}
then convert this from time slots to flights, i.e.\ $\modMatch[{i,j}]$ is the
flight in which skippers $s[{i}]$ and $s[{j}]$ meet for a match. Finally,
constraint \ref{eq19} ensures that each skipper's match occurs in different
flights. Also, since $\modMatch[{i,j}] \equiv \modMatch[{j,i}]$, setting rows
to be all different also ensures that columns are all different.

%
% the match array
%
\begin{flalign}
    & \forall_{i \in [0..n-2]} ~ \forall_{j \in [i+1..n-1]}  ~\match[{i,j}] \in \{0..t-1\} \label{eq13} \\
    & \forall_{i \in [0..n-2]} ~ \forall_{j \in [i+1..n-1]}  ~\match[{j,i}] \equiv \match[{i,j}] \label{eq14} \\
    & \forall_{i \in [0..n-2]} ~ \forall_{j \in [i+1..n-1]}  ~ \forall_{k \in [0..t-1]} ~ \match[{i,j}] = k \iff s[{i}].\timeSlot[{k}] = j  ~ \wedge ~ s[{j}].\timeSlot[{k}] = i  \label{eq15} \\
    & \forall_{i \in [0..n-2]} ~ \forall_{j \in [i+1..n-1]}  ~ \modMatch[{i,j}] \in \{0..f-1\} \label{eq16} \\
    & \forall_{i \in [0..n-2]} ~ \forall_{j \in [i+1..n-1]}  ~ \modMatch[{j,i}] \equiv \modMatch[{i,j}] \label{eq17} \\
    & \forall_{i \in [0..n-2]} ~ \forall_{j \in [i+1..n-1]}  ~ \modMatch[{i,j}] = \frac{\match[{i,j}]}{m} \label{eq18} \\
    & \forall_{i \in [0..n-1]} ~ \allDifferent(\modMatch[{i}]) \label{eq19}
\end{flalign}

We now take a temporal view (\ref{eq20}), such that $\timeVar[{\tau}]$ is a pair
$(i,j)$, stating that at time $\tau$ skippers $s[{i}]$ and $s[{j}]$ are in a
match, i.e.\ we channel between the match-perspective and the
temporal-perspective.
%
% one dimensional time and decision variables
%
\begin{flalign}
    & \forall_{\tau \in [0..t-1]} ~ \timeVar[{\tau}] \in \{(0,1) .. (n-2,n-1)\} \label{eq20} \\
    & \forall_{\tau \in [0 .. t-1]} ~ \forall_{i \in [0..n-2]} ~ \forall_{j \in [i+1..n-1]}  ~ \timeVar[{\tau}] = (i,j) \iff \match[{i,j}] = \tau \label{eq21}
\end{flalign}

Finally we have the optimisation criteria (criteria 10) to minimise the number
of boat changes. This is presented in \ref{eq22} and \ref{eq23} below.

%
% optimisation criteria 10
%
\begin{flalign}
    & \totalBoatChanges = \sum_{i=0}^{n-1} s[{i}].\totalChanges  \label{eq22} \\
& \minimise(\totalBoatChanges) \label{eq23}
\end{flalign}

>>>>

The first thing we model, in the first half of Figure~\ref{model:stage1}, is a
skipper. Each skipper $\sigma$ has a temporal view of his schedule in the array
of constrained integer variables $\timeSlot[0]$ to $\timeSlot[{t-1}]$, and a
state view in the array of constraint integer variables $\state[0]$ to
$\state[{f-1}]$. The state of this skipper in flight $i$ is found in
$\state[i]$, and this corresponds to timeslots $\timeSlot[{i \cdot m}]$ to
$\timeSlot[{m(i+1)-1}]$.

\begin{figure}[p]
\setlength{\mathindent}{1em}
\setlength{\abovedisplayskip}{0pt}
\setlength{\belowdisplayskip}{0pt}
\setlength{\abovecaptionskip}{0pt}
\begin{framed}
\begin{align}
    \intertext{A copy of these variables and constraints is created for each skipper $\sigma$:\smallskip}
    &\forall \tau \in \{0 \nldots t{-}1\}: \timeSlot[{\tau}] \in \{{-}2 \nldots t{-}1\} \setminus \{\sigma\} \label{var:timeSlot} \tag{V1} \\
    &\forall i \in \{0 \nldots f{-}1\}: \state[{i}] \in \{\FIRST,\MID,\LAST,\BYE,\END\} \label{var:state} \tag{V2} \\
    &\forall i \in \{0 \nldots f\}: \label{C1} \tag{C1} \\
    &\begin{alignedat}{3}
    &\hspace*{1em} \state[{i}] = \FIRST &\ &\Leftrightarrow&&\ \timeSlot[{i \cdot m}] \geq 0 \\
    &\hspace*{1em} \state[{i}] = \MID   &\ &\Leftrightarrow&&\ \exists j \in \{i\cdot m{+}1 \nldots i \cdot m{+}m{-}2\} : \timeSlot[{j}] \geq 0 \\
    &\hspace*{1em} \state[{i}] = \LAST  &\ &\Leftrightarrow&&\ \timeSlot[{(i{+}1)m {-} 1}] \geq 0 \\
    &\hspace*{1em} \state[{i}] = \BYE   &\ &\Leftrightarrow&&\ \forall j \in \{i \cdot m \nldots m \cdot (i{+}1){-}1\} : \timeSlot[j] = {-}1 \\
    &\hspace*{1em} \state[{i}] = \END   &\ &\Leftrightarrow&&\ \forall j \in \{i \cdot m \nldots m \cdot (i{+}1){-}1\} : \timeSlot[j] = {-}2
    \end{alignedat} \nonumber \\
    &\eachOccursExactlyOnce(\timeSlot, \{0, \nldots, n{-}1\} \setminus \{\sigma\}) \label{C2} \tag{C2} \\
    &\validStateSequence(\state) \label{C3} \tag{C3} \\[0.1cm]
    & \forall i \in \{0 \nldots f{-}2\}: \change[{i}] \in \{0,1\} \label{V3} \tag{V3} \\
    & \totalChanges \in \mathbb{N} \label{V4} \tag{V4} \\
    & \forall i \in \{0 \nldots f{-}2\}: \change[{i}] = 1 \Leftrightarrow \state[{i}] = \BYE \wedge \state[{i{+}1}] \neq \BYE \label {C4} \tag{C4} \\
    & \totalChanges = \textstyle\sum_{i=0}^{f{-}2} \change[{i}] \label{C5} \tag{C5}
\end{align}
\end{framed}\begin{framed}
\begin{align}
    \intertext{Match and temporal perspectives:\smallskip}
    & \forall i \in \{0 \nldots n{-}2\} : \forall j \in \{ i{+}1 \nldots n{-}1 \} : \nonumber \\
    & \hspace*{1em} \match[{i,j}] \in \{0..t{-}1\} \tag{V5} \label{V5} \\
    & \hspace*{1em} \match[{j,i}] \equiv \match[{i,j}] \nonumber \\
    & \hspace*{1em} \forall k \in \{0 \nldots t{-}1 \} : \nonumber \\
    & \hspace*{2em} \match[{i,j}] = k \Leftrightarrow \sigma[{i}].\timeSlot[{k}] = j \wedge \sigma[{j}].\timeSlot[{k}] = i \tag{C6} \label{C6} \\[0.1cm]
    & \forall i \in \{0 \nldots n{-}2\} : \forall j \in \{ i{+}1 \nldots n{-}1 \} \nonumber : \\
    & \hspace*{1em} \modMatch[{i,j}] \in \{0..f{-}1\} \label{V6} \tag{V6} \\
    & \hspace*{1em} \modMatch[{j,i}] \equiv \modMatch[{i,j}] \nonumber  \\
    & \hspace*{1em} \modMatch[{i,j}] = \match[{i,j}] / m \tag{C7} \label{C7} \\
    & \forall i \in \{0 \nldots n{-}1 \} : \allDifferent(\modMatch[{i}]) \tag{C8} \label{C8} \\[0.1cm]
    & \forall \tau \in \{0 \nldots t{-}1 \} : \nonumber \\
    & \hspace*{1em} \timeVar[{\tau}] \in \{(0,1) \nldots (n{-}2,n{-}1)\} \tag{V7} \label{V7} \\
    & \hspace*{1em} \forall i \in \{0 \nldots n{-}2\} : \forall  j \in \{ i{+}1 \nldots n{-}1 \} : \timeVar[{\tau}] = (i,j) \Leftrightarrow \match[{i,j}] = \tau \tag{C9} \label{C9} \\[0.1cm]
    & \totalBoatChanges = \textstyle\sum_{i=0}^{n{-}1} \sigma[{i}].\totalChanges  \label{V8} \tag{V8} \\
    & \minimise(\totalBoatChanges) \label{C10} \tag{C10}
\end{align}
\end{framed}
\caption{Our constraint model, from a skipper perspective (top) and a match and
temporal perspective (below). The number of skippers is $n$, and $m$ is the
number of matches in a flight. The number of flights is $f = \lceil n(n-1)/m
\rceil$, and there are $t = n(n-1)/2$ matches in total.}\label{model:stage1}
\end{figure}

@@

We are now in a position to model criteria 1 (each skipper sails against every
other skipper) and optimisation criteria 10 (minimise boat changes).  In the
second half of Figure~\ref{model:stage1} we present a match perspective of the
schedule, using a two dimensional array of variables $\match$ \eqref{V5}, where
$\match[{i,j}]$ is the time slot that skippers $\sigma[{i}]$ and $\sigma[{j}]$
meet in a match. Only the half above the diagonal is represented, and the lower
half is made up of exactly the same variables, i.e.\ $\match[{i,j}]$ is
\emph{exactly} the same constrained variable as $\match[{j,i}]$. Constraint
\eqref{C6} states that  a match between skippers $\sigma[{i}]$ and
$\sigma[{j}]$ takes place at time $k$ (i.e.\ $\match[{i,j}] = k$) if and only
if  skipper $\sigma[{i}]$'s $k^{th}$ time slot is skipper $\sigma[{j}]$ and
conversely that skipper $\sigma[{j}]$'s $k^{th}$ time slot is skipper
$\sigma[{i}]$). Variables \eqref{V6} and constraints \eqref{C7} then convert
this from time slots to flights, i.e.\ $\modMatch[{i,j}]$ is the flight in
which skippers $\sigma[{i}]$ and $\sigma[{j}]$ meet for a match. Finally,
constraint \eqref{C8} ensures that each skipper's match occurs in different
flights. Also, since $\modMatch[{i,j}] \equiv \modMatch[{j,i}]$, setting rows
to be all different also ensures that columns are all different.

We now take a temporal view \eqref{V7}, such that $\timeVar[{\tau}]$ is a pair
$(i,j)$, stating that at time $\tau$ skippers $\sigma[{i}]$ and $\sigma[{j}]$
are in a match. We channel between the match perspective and the
temporal perspective using \eqref{C9}.

Finally we have the optimisation criteria (criteria 10) to minimise the number
of boat changes. This is handled by \eqref{V8} and \eqref{C10}.
